A Tale of Two Dictionary Learning Problems

### 2020-08-14 manuscripts/thesis/abstract.tex
almost done

Two formulations of the dictionary learning problem are considered. 
The first seeks a dictionary for which each signal in a dataset is well-approximated by a linear superposition of a limited number of dictionary elements.
%he first seeks a dictionary admitting a sparse representation of limited support size for each signal in a dataset. 
%In this case, beyond an original extension of existing guarantees to the noisy regime, a theory of combinatorial designs is introduced to show that the dictionary and corresponding sparse codes are almost always identifiable up to an error commensurate with the noise even if the dictionary fails to satisfy the spark condition, its size is overestimated, or it induces only a polynomial number of distinct sparse supports in the data. 
In this case, existing guarantees are extended to the noisy regime to demonstrate that such dictionaries and the sparse representations they induce are almost always identifiable up to an error commensurate with the noise. Moreover, a theory of combinatorial designs is introduced to demonstrate this is the case even if the dictionary fails to satisfy the spark condition, the data are distributed over only a polynomial set of subspaces spanned by the dictionary, or (to some extent) even if the dictionary is overestimated in size.

@@

This work serves to justify, in principle, dictionary learning in general as a means of discovering latent sparse structure in real data. Though much work remains to be done deriving criteria for use in practice, the theoretical tools developed herein should be of use to this end.
%The work closes with some open questions and directions for future research, seeded in part by the results of more practically-minded simulations. 
\end{abstract}

>>>>

Two formulations of the dictionary learning problem are considered. 
The first seeks a dictionary for which each signal in a dataset is well-approximated by a linear superposition of only a limited number of dictionary elements.
%he first seeks a dictionary admitting a sparse representation of limited support size for each signal in a dataset. 
%In this case, beyond an original extension of existing guarantees to the noisy regime, a theory of combinatorial designs is introduced to show that the dictionary and corresponding sparse codes are almost always identifiable up to an error commensurate with the noise even if the dictionary fails to satisfy the spark condition, its size is overestimated, or it induces only a polynomial number of distinct sparse supports in the data. 
In this case, existing guarantees are extended to the noisy regime to demonstrate that such dictionaries and the sparse representations they induce are almost always identifiable up to an error commensurate with the noise. Moreover, a theory of combinatorial designs is introduced to demonstrate this is the case even if the dictionary fails to satisfy the spark condition, the data are distributed over only a polynomial set of subspaces spanned by the dictionary, or (to some extent) even if the dictionary is overestimated in size.

@@

This work serves to justify, in principle, dictionary learning in general as a means of discovering latent sparse structure in real data. Though much work remains to be done deriving criteria for use in practice, the theoretical tools developed here should be of use to this end.
%The work closes with some open questions and directions for future research, seeded in part by the results of more practically-minded simulations. 
\end{abstract}

### 2020-08-14 manuscripts/thesis/discussion.tex
almost done

\subsection{Signal-to-Noise Ratio}

@@

\subsection{Sample Complexity}

@@

A close examination of how exactly Lemma \ref{NonEmptyLemma} is incorporated into the proof of Lem.~\ref{MainLemma} reveals, however, that looser constraints on $\mathcal{H}$ maintain compatibility with the argument to some degree. Assuming that the nodes of $\mathcal{H}$ are labeled in some order of non-increasing degree, if instead of the full SIP we require only that the stars $\cap \sigma(i)$ intersect at singletons for all $i \leq q$ for some positive $q$, we have that $\overline m \geq k|\mathcal{H}| / \deg(1)$ and, provided $\overline m < k|\mathcal{H}| / (\deg(1) - 1)$, the nonempty subset $J$ is of size equal to the largest number $p$ satisfying:
\begin{align*}
\sum_{i=\ell}^{m} \deg(i) > (\overline m + 1 - \ell) (\deg(\ell) - 1) \ \ \text{for all } \ell \leq p \leq q.
\end{align*}
Specifically, $J$ contains all nodes of degree exceeding $\deg(p)$ and some subset of those with degree equal to $\deg(p)$. It is thus natural to ask: what are the necessary constraints on the hypergraph $\mathcal{H}$, and what is the smallest hypergraph satisfying these constraints for given $m$ and $k$?
%%%%%%%%%%%%%%%%

@@

We may also elaborate on the probabilistic pigeonholing strategy outlined in the discussion following the proof of Thm.~\ref{DeterministicUniquenessTheorem}. The problem is to count the number of ways in which vectors supported in $S \in \mathcal{H}$ can be partitioned among supports in ${[\overline m] \choose k}$ without allocating $k$ or more to any individual one (in which case the logic of the proof fails to imply the result; we are interested in the probability that it doesn't). These are integer solutions to the problem $\sum_i n_i = N$ subject to $n_i < k$ for all $i$, where $i = 1, \ldots, {\overline m \choose k}$. Following closely the exposition in \cite{stackexchangeanswer}, it appears there is no closed formula for this problem, but the number of solutions can be computed in a number of operations independent of $N$. Writing $p = {\overline m \choose k}$, the number is the coefficient of $X^N$ in the polynomial $(1 + X + \ldots + X^{k-1})^p$. Written as a rational function of $X$,

@@

\subsection{Dictionary learning via convex optimization}

>>>>

\subsection{Reducing the required signal-to-noise ratio}

@@

\subsection{Reducing the required number of samples}

@@

A close examination of the proof of Lemma \ref{NonEmptyLemma} and its incorporation into the proof of Lem.~\ref{MainLemma} reveals, however, that looser constraints on $\mathcal{H}$ maintain compatibility with the argument to some degree. Assuming that the nodes of $\mathcal{H}$ are labeled in order of non-increasing degree, if instead of the full SIP we require only that the stars $\cap \sigma(i)$ intersect at singletons for all $i \leq q$ for some positive $q$, we have that $\overline m \geq k|\mathcal{H}| / \deg(1)$ and, provided $\overline m < k|\mathcal{H}| / (\deg(1) - 1)$, the nonempty subset $J$ is of size equal to the largest number $p$ satisfying:
\begin{align*}
\sum_{i=\ell}^{m} \deg(i) > (\overline m + 1 - \ell) (\deg(\ell) - 1) \ \ \text{for all } \ell \leq p \leq q.
\end{align*}
Specifically, $J$ contains all nodes of degree exceeding $\deg(p)$ and some subset of those with degree equal to $\deg(p)$. It is thus natural to ask: what are the necessary constraints on the hypergraph $\mathcal{H}$, and what is the smallest hypergraph satisfying these constraints for given $m$ and $k$?
%%%%%%%%%%%%%%%%

@@

We may also elaborate on the probabilistic pigeonholing strategy outlined in Sec.~\ref{DUT} following the proof of Thm.~\ref{DeterministicUniquenessTheorem}. The problem is to count the number of ways in which vectors supported in $S \in \mathcal{H}$ can be partitioned among supports in ${[\overline m] \choose k}$ without allocating $k$ or more to any individual one (in which case the logic of the proof fails to imply the result; we are interested in the probability that it doesn't). These are integer solutions to the problem $\sum_i n_i = N$ subject to $n_i < k$ for all $i$, where $i = 1, \ldots, {\overline m \choose k}$. Following closely the exposition in \cite{stackexchangeanswer}, it appears there is no closed formula for this problem, but the number of solutions can be computed in a number of operations independent of $N$. Writing $p = {\overline m \choose k}$, the number is the coefficient of $X^N$ in the polynomial $(1 + X + \ldots + X^{k-1})^p$. Written as a rational function of $X$,

@@

\subsection{Dictionary learning via $\ell_1$-norm minimization}

### 2020-08-14 manuscripts/thesis/thesis.tex
almost done

\begin{dedication}
\null\vfil
%\begin{center} 
I want to dedicate this work to my dad. At the same time, I don't want to, because these results can't be matched to the time with him I traded to produce them, and it feels as though dedicating this work to him would in some sense be measuring his life up against them.

The fact is that whatever may be the eventual butterfly effect of this work on the scientific ecosystem, these mathematical truths could have waited, arrived at their own pace, via myself or another, and I could have been more present with my dad as he waited patiently to die of cancer.

I suppose that I couldn't really fathom what it actually meant for this to happen -- to never see or talk to or hug him again -- while the idea of putting original work out there, of crafting my own piece of the overwhelming puzzle that is neuroscience, was tantalizing. Yet it was more-so a feeling of desperation, of having to accomplish something and prove myself worthy of the position I had come to find myself in after 25+ years of schooling and a first failed project of my own design, that I feel largely drove these efforts. Truth be told, I initially had no particular interest in dictionary learning other than as a means to passing my qualifying exam; and by that time, I worried the brain was so dauntingly complex that I should feel lucky to have at least been offered a well-defined and arguably relevant problem in neuroscience to work on, particularly one I felt even remotely equipped to solve myself.

Let's say, then, that this is the first bit of a life's work dedicated to my dad, real work devoid of any prestigious position and the associated self-doubt, pressure, or delusions of grandeur; a life dedicated to passionate ideas, to solving problems I care about and that really matter, to loving and caring for others or, as my dad would say, to ``being a mensch". May this dedication serve to hold me publicly accountable to that. Also I love you, mom. I dedicate and don't dedicate this and everything to come to you, too.

>>>>

\begin{dedication}
\null\vfil
%\begin{center} 
I want to dedicate this work to my dad. At the same time, I don't want to, because these results can't be matched to the time with him I traded to produce them, and it feels as though dedicating this work to him would in some sense be measuring his life up against them. The fact is that whatever may be the eventual butterfly effect of this work on the scientific ecosystem, these mathematical truths could have waited, arrived at their own pace, via myself or another, and I could have been more present with my dad as he waited patiently to die of cancer.

I suppose that I couldn't really fathom what it actually meant for this to happen -- to never see or talk to or hug him again -- while the idea of putting original work out there, of crafting my own piece of the overwhelming puzzle that is neuroscience, was tantalizing. Yet it was more-so a feeling of desperation, of having to accomplish something and prove myself worthy of the position I had come to find myself in after 25+ years of schooling and a first failed project of my own design, that I feel largely drove these efforts. Truth be told, I initially had no particular interest in dictionary learning other than as a means to passing my qualifying exam; and by that time, I worried the brain was so dauntingly complex that I should feel lucky to have at least been offered a well-defined problem of arguable relevance to neuroscience to work on, particularly one I felt even remotely equipped to solve myself.

Let's say, then, that this is the first bit of a life's work dedicated to my dad, work devoid of any prestigious position and the associated self-doubt, pressure, or delusions of grandeur; a life dedicated to passionate ideas, to solving problems I actually care about and that really matter, to loving and caring for others or, as my dad would say, to ``being a mensch". May this dedication serve to hold me publicly accountable to that. Also I love you, mom. I dedicate and don't dedicate this and everything to come to you, too.
